Probabilistic Catalogs for Crowded Stellar Fields

### 2013-02-28 paper/ms.tex
Went through first half of paper, hopefully clearing up exposition. I actually thought it read well already

\begin{abstract}
We present a probabilistic (Bayesian) method for producing catalogs
from images of stellar fields. The method is capable of inferring the
number of sources $N$ in the image and can also handle the
challenges introduced by noise and overlapping sources. The luminosity
function of the stars can also be inferred even when the precise luminosity
of each star is uncertain, via the use of a hierarchical Bayesian model.
The computational feasibility of the method is
demonstrated on two simulated images with different numbers of stars.
We find that our method
successfully recovers the input parameter values along with principled
uncertainties. We also compare our results with those obtained from the
\sex~software and find that the latter underestimates the number of stars in the
crowded case and ...
\end{abstract}

@@

Telescopes don't make catalogs \citep{2011EAS....45..351H}, they
measure the intensity field.  Viewed through the lens of probabilistic
inference, the goals of astronomy are to take the information in the
telescope-generated records of the intensity field and deliver it to
the quantities of astronomical interest with as little loss as
possible.  Insertion of a catalog-generation step in the inference
pipeline between the raw imaging and the final astrophysical analyses
is potentially very lossy; the hard decisions of catalog making destroy
information.  Probability theory suggests that it may be less
lossy to pass forward not a catalog but a probabilisitic
description of all the catalogs that could be consistent with the
imaging---a posterior probability in (enormously large) catalog space.
This article represents an attempt at this ambitous goal in the specific
situation where the only objects in the field are stars or other point sources.

Beyond these extremely general concerns, there are practical issues:
Standard methods for constructing catalogs can have difficulty in some
challenging situations. For example, when multiple sources overlap partially
or completely, it can be difficult to determine how many sources are present,
and how much flux belongs to each source. In principle, the uncertainty about
the existence and properties of the stars can be significant and should be
propagated into any inferences about the stellar population.

Essentially, the creation of a catalog is an
attempt to answer the question,
``Given the image we have obtained, what objects are present in the field and
what are their properties?'' This motivates a probabilistic (Bayesian)
approach to making catalogs. The output of such an approach would not be a
single answer to this question (i.e. a single ``point estimate'' catalog), but
rather a posterior probability distribution over the space of possible catalogs.
Numerically, Bayesian Inference is often implemented using Markov Chain Monte
Carlo (MCMC) algorithms in order to generate random samples from the posterior
distribution.

Sampling a posterior probability distribution for catalogs is likely to be
challenging for a number of reasons. Firstly, the number $N$ of objects in the
image (and that should therefore be listed in the catalog) is itself unknown.
Secondly, if $N$ is large, then the parameter space
of positions and properties (flux, size, etc) of the objects is also large.
This can cause Markov Chain Monte Carlo (MCMC) algorithms difficulties -- they
may take a long time to converge to the target posterior distribution over
the space of catalogs. Thirdly, this problem is subject to the so-called
label-switching problem that is commonly encountered in mixture modeling
\citep[e.g.][]{label_switching}. Given any proposed catalog, another catalog that is
equally plausible is the catalog obtained by shuffling the entries of the first
catalog. This leads to a posterior distribution with $N!$ identical peaks in
parameter space. This can lead to difficulties with certain (generally very
effective) MCMC algorithms such as the affine-invariant stretch move
\citep{goodman, emcee}.

Bayesian object detection (as this problem is sometimes called) has been
implemented both inside and outside of astronomy
\citep[e.g.][]{object, 2003MNRAS.338..765H, 2011MNRAS.415.3462F}. However, the
\citet{2011MNRAS.415.3462F} approach makes the assumption of a
known number of objects $N$.
This assumption is required for the
{\tt MultiNest} sampler \citep{multinest} to be applicable.
Using the results from the known $N$
run, it is possible (under certain circumstances) to reconstruct what the
results would have been if an unknown-$N$ model had been used. However,
this will not work well in situations where there is significant confusion
(i.e. two or more sources overlap). What is really required is a variable
dimension model, where $N \in \{0, 1, 2, ... \}$ is an unknown quantity to be
inferred from the data.
The computational implementation of these models will require tools such as
reversible jump Markov Chain Monte Carlo \citep{rjmcmc}. Other statistical
methods have also been used to model crowded fields
\citep[e.g. maximum likelihood,][]{irwin}.
However, maximum likelihood is not completely appropriate in flexible models
because it may lead to overfitting. In this situation overfitting would result
in more stars being added to the model to explain small positive fluctuations
in the image which are actually due to noise. Various other techniques have
also been proposed in the literature \citep[e.g.][]{2002MNRAS.335...73M, 2007A&A...461..373M, 2009PASP..121.1429Z}.

@@

\section{Bayesian Inference}\label{sec:bayes}
To quantitatively model uncertainties and transform noise in observed data
into uncertainties in parameters of interest, Bayesian Inference is the appropriate
framework \citep{cox, jaynes, caticha, mackay}. Suppose there exist unknown parameters
(denoted
collectively by $\theta$) and we expect to obtain some data $x$. Our prior
state of knowledge about the parameters is modelled by a prior
probability distribution:
\begin{equation}
p(\theta).
\end{equation}
Note that this is a very concise notation \citep{hogg} and should be read
as ``the probability distribution for $\theta$''.
We also model how the parameters give rise to the data, via a generative model.
This is also known as a {\it sampling distribution}:
\begin{equation}
p(x|\theta).
\end{equation}
Despite the singular, the sampling distribution is actually a family of
probability distributions over the space of possible data sets, one probability
distribution for each possible value
of $\theta$. Note that the choice of
the sampling distribution is also an assumption about prior knowledge:
It models prior information about the fact that the data $x$ is connected to
the parameters $\theta$ in some way \citep{caticha}. Without this prior
knowledge, learning is impossible: there has to be some relationship between
the parameters and the data, otherwise new data tells you about nothing but
itself.

When specific data $x^*$ have been obtained, our state of knowledge about $\theta$
gets updated from the prior distribution to the posterior distribution
via Bayes' rule:
\begin{eqnarray}
p(\theta|x=x^*) &\propto& p(\theta)p(x|\theta)|_{x=x^*} \\
&=& p(\theta)\mathcal{L}(\theta; x)
\end{eqnarray}
The term $p(x|\theta)|_{x=x^*} = \mathcal{L}(\theta; x)$
is the {\it likelihood function}, which is the
probability of obtaining the actual data set $x^*$ as a function of the
parameters. In the case that the sampling distribution is a probability
density, the likelihood is the probability density function evaluated at the observed
data. This usually causes no problems, although one should be aware of the
Borel-Kolmogorov paradox \citep{jaynes}.
As suggested by the above notation, the likelihood function is obtained from the
sampling distribution with the actual data substituted in and is therefore
a function of the parameters only.

@@

For simplicity, we assume a uniform prior probability distribution for the
position of each star. This creates a strong preference for catalogs where
the stars are uniformly distributed across the image. Thus, this model is
appropriate for small patches of sky where the density of stars is approximately
uniform. In other scenarios, such as images of stellar clusters, it is possible
to parameterize the spatial distribution of the stars in a similar way to how
we have parameterized the luminosity function.

\subsection{The Sampling Distribution}
The sampling distribution is a probabilistic model for the process that
generates the data; it describes the probability distribution we would use
to predict the data if we happened to know the true catalog. In our case,
the data will be an $m \times n$ array of pixel intensities $I$:
\begin{eqnarray}
\{I_{ij}\}
\end{eqnarray}
where the central position of each pixel is:
\begin{eqnarray}
\{X_{ij}, Y_{ij}\}.
\end{eqnarray}
The image is assumed to be a noisy version of the true underlying intensity
field. Thus, we need
a prescription for simulating an image $\{I_{ij}\}$ from a catalog $\theta$:
\begin{eqnarray}
\theta = \left\{N, \beta, \left\{x_i, y_i\right\}_{i=1}^N,
\left\{f_i\right\}_{i=1}^N\right\}.
\end{eqnarray}
If we knew the true catalog, we could compute the
``mock image'' we would expect to see
in the absence of noise. This mock image (defined at every point on the sky)
is given by:
\begin{eqnarray}
\mathcal{M}(x, y) &=& \sum_{i=1}^N f_i \mathcal{P}(x - x_i, y - y_i)
\end{eqnarray}
where $\mathcal{P}$ is the pixel-convolved point spread function (PSF).
Throughout this paper we will
assume the pixel-convolved PSF is a weighted mixture of two concentric
circular
Gaussians with widths $s_1$ and $s_2$:
\begin{eqnarray}
\mathcal{P}(x, y) = \frac{w}{2\pi s_1^2}\exp
\left[
-\frac{1}{2s_1^2}\left(x^2 + y^2\right)
\right]
+ \frac{1-w}{2\pi s_2^2}\exp
\left[-\frac{1}{2s_2^2}\left(x^2 + y^2\right)
\right].
\end{eqnarray}
The pixellated observed image is assumed to be generated from the mock
image (evaluated at the center of each pixel) plus noise:
\begin{eqnarray}
I_{ij} &=& \mathcal{M}(X_{ij}, Y_{ij}) + \epsilon_{ij}
\end{eqnarray}
where the errors $\{\epsilon_{ij}\}$ are independent and normally distributed.
The variance of each pixel is determined by the brightness of the sky and the
brightness of the mock image in that pixel:
\begin{eqnarray}
\epsilon_{ij} \sim \mathcal{N}(0, \sigma_0^2 + \eta \mathcal{M}(X_{ij}, Y_{ij})).
\end{eqnarray}
where $\sigma_0$ is a constant noise level and $\eta$ is a coefficient allowing
the noise to be higher in brighter regions of the image. This dependence of the
noise variances on the model intensity arises as a result of the Poissonian
nature of photon counts, but allows for the fact that a ``sky'' background may
have already been subtracted from the image.
This parameterisation has been used by \citet{2011MNRAS.412.2521B} and is an
alternative to the common practice of producing a ``variance map'' from the
image data that is then assumed to be known.

\subsection{The Prior Distribution}
The prior distribution for the number of stars $N$ is assigned to be uniform
between 0 and some maximum number $N_{\rm max}$. The extent of the image is
assumed to be from $x=x_{\rm min}$ to $x=x_{\rm max}=x_{\rm min} + x_{\rm range}$ and from
$y=y_{\rm min}$ to $y=y_{\rm max}=y_{\rm min} + y_{\rm range}$ in arbitrary units, and the
positions of the stars are assigned independent uniform priors:
\begin{eqnarray}
x_i &\sim& \textnormal{Uniform}(x_{\rm min} - 0.1x_{\rm range}, x_{\rm max} + 0.1x_{\rm range}) \\
y_i &\sim& \textnormal{Uniform}(y_{\rm min} - 0.1y_{\rm range}, y_{\rm max} + 0.1y_{\rm range})
\end{eqnarray}
The stars are allowed to be slightly outside of the observed image because the
PSF can scatter light from these stars into the image. Our model for the
luminosity function is a broken power-law with four
free parameters:
\begin{eqnarray}
\beta = \{h_1, h_2, \alpha_1, \alpha_2\}.
\end{eqnarray}
where $h_1$ is a lower flux limit, $h_2$ is a break-point, $\alpha_1$ is
the slope of the distribution between $h_1$ and $h_2$, and $\alpha_2$ is the
slope of the distribution above $h_2$. For details on the broken power-law
model, see Appendix~\ref{power_law}. The prior distribution on $h_1$, $h_2$,
$\alpha_1$, and $\alpha_2$ is assigned to be:
\begin{eqnarray}
\ln h_1 &\sim& \textnormal{Uniform}(\ln(10^{-3}), \ln(10^3)) \\
\ln h_2 &\sim& \textnormal{Uniform}(\ln(h_1), \ln(h_1) + 2.3) \\
\alpha_1 &\sim& \textnormal{Uniform}(1, 5) \\
\alpha_2 &\sim& \textnormal{Uniform}(1, 5).
\end{eqnarray}
These priors express vague prior knowledge about $\alpha_1$ and $\alpha_2$ in
addition to vague prior knowledge about $h_1$ and $h_2$ apart from the fact
that the flux units are not extreme and that $h_2$ should be no more than an
order of magnitude greater than $h_1$.

@@

The priors for the PSF parameters and the noise parameters were:
\begin{eqnarray}
\ln s_1 &\sim& \textnormal{Uniform}(\ln(0.3L), \ln(30L)) \\
\ln s_2 &\sim& \textnormal{Uniform}(\ln(s_1), \ln(s_1) + 2.3) \\
w &\sim& \textnormal{Uniform}(0, 1)\\
\ln \sigma_0 &\sim& \textnormal{Uniform}(\ln(10^{-3}), \ln(10^{3})) \\
\ln \eta &\sim& \textnormal{Uniform}(\ln(10^{-3}), \ln(10^{3}))
\end{eqnarray}
where $L = (x_{\rm max} - x_{\rm min})/n$ is the width of a pixel.

@@

\section{Simulated Data}\label{sec:simulated_data}
In order to test our approach, we applied the method to two illustrative
simulated images generated from
the above model. The purpose of this experiment was to test the computational
feasibility of the model, as well as to compare the inferences from the model
with those from more standard techniques.

The true parameter values for the two simulated data sets are listed in
Table~\ref{tab:truth}. The broken power-law parameter values were chosen so that
roughly half of the stars' fluxes were below and above the break-point
respectively. Figure~\ref{fig:powerlaw} in Appendix~\ref{power_law} also shows the true flux distribution
used for the simulated images. Each of the images
is $100 \times 100$ pixels in extent and covers a range from $-1$ to $1$ in
arbitrary units for both the $x$ and $y$ axes. The first image contains
100 stars (including stars just outside of the image; there are 63 stars actually
centered within the image) and the second
image contains $\sim$1000 stars (699 of which are centered within the image).
The two simulated images themselves
are shown in Figure~\ref{fig:simulated_data}.

\begin{table}\footnotesize
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Parameter & Value (Test Case 1) & Value (Test Case 2)\\
\hline
$N$ & 100 & 1000\\
$h_1$ & 0.3 & 0.3\\
$h_2$ & 0.6 & 0.6\\
$\alpha_1$ & 1.1 & 1.1\\
$\alpha_2$ & 2 & 2\\
\hline
$\sigma_0$ & 10 & 10\\
$\eta$ & 10 & 10 \\
$s_1$ & 0.02 & 0.02\\
$s_2$ & 0.1 & 0.1\\
$w$ & 0.5 & 0.5\\
\hline
\end{tabular}
\end{center}
\caption{True parameter values used to generate the simulated data. The main
difference between the two test cases is the number of stars and the noise
level, both of which are higher in test case 2.
\label{tab:truth}}
\end{table}

@@

\subsection{Test Case 1}
Test Case 1 was run with the DNS algorithm and usable results were obtained
within about an hour on a modern desktop PC. The inferences on the parameters
$N$, $h_1$, $h_2$, $\alpha_1$, and $\alpha_2$ are shown in
Figure~\ref{fig:results1}. The number of stars is correctly inferred, and the
posterior distributions for the other parameters comfortably contain the true
input values. Note that the uncertainty in $h_2$, $\alpha_1$, and $\alpha_2$ is
quite large. This is because the broken power-law model
(Figure~\ref{fig:powerlaw}) does not change drastically in shape as the
parameters are varied. Therefore, a large number of stars would be required to
tightly constrain the parameters. A Nested Sampling approach assists on
this problem because of the presence of a first-order phase transition
\citep{skilling}.

@@

\subsection{Test Case 2}
Test Case 2 is more challenging than Test Case 1 because the image contains
more stars. This increases the size of the computational task in a number of
ways: firstly, there will be more unknown parameters to infer, so the
convergence of the MCMC algorithm will take longer. Secondly, the time taken
to compute the predicted image from a catalog (in order to evaluate the
likelihood) is longer because of the larger number of stars. Using DNS, some
samples from the posterior distribution can be obtained in about a day on a
modern PC. However, this problem does not have a first-order phase transition
like Test Case 1. Therefore, it is tractable to run standard Metropolis-Hastings
MCMC targeting the posterior distribution. In fact, this can be more efficient
because the efficiency of Nested Sampling methods is sensitive to the ``information'',
or the amount of prior mass occupied by the posterior distribution. In this case,
because of the large number of parameters, the information is very large, about
3000 nats.

@@

\begin{table}\footnotesize
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Parameter & Proposal & Notes\\
\hline
$N$ & $N \to N + \delta_N$ & Generate $\delta_N$ new stars from
$p(x, y, f|\beta)$.\\
$N$ & $N \to N - \delta_N$ & Remove $\delta_N$ stars, chosen at random\\
$\beta$ & $\beta \to \beta + \delta_\beta$ & Transform
stars' fluxes correspondingly\\
$\beta$ & $\beta \to \beta + \delta_\beta$ & Fix stars' fluxes, put extra
term in acceptance probability \\
$(x_i,y_i)$ & $(x_i,y_i) \to (x_i,y_i)+(\delta_x, \delta_y)$ & Can move $>1$ star
in a single step \\
$f$ & $f \to f + \delta_f$ & Can move $>1$ stars' fluxes in a single step\\
\hline
\end{tabular}
\end{center}
\caption{All $\delta$ parameters are drawn from multi-scale distibutions such
that the largest steps are of order the prior width, and the smallest steps
are of order $10^{-6}$ times the prior width.\label{tab:proposals}}
\end{table}

>>>>

\begin{abstract}
We present a probabilistic (Bayesian) method for producing catalogs
from images of stellar fields. The method is capable of inferring the
number of sources $N$ in the image and can also handle the
challenges introduced by noise and overlapping sources. The luminosity
function of the stars can also be inferred even when the precise luminosity
of each star is uncertain, via the use of a hierarchical Bayesian model.
The computational feasibility of the method is
demonstrated on two simulated images with different numbers of stars.
We find that our method
successfully recovers the input parameter values along with principled
uncertainties even when the field is crowded.
We also compare our results with those obtained from the
\sex~software and find that the latter underestimates the number of stars in the
crowded case and ...
\end{abstract}

@@

Telescopes don't make catalogs \citep{2011EAS....45..351H}, they
measure the intensity field.  Viewed through the lens of probabilistic
inference, the goals of astronomy are to take the information in the
telescope-generated records of the intensity field and deliver it to
the quantities of astronomical interest with as little loss as
possible.  Insertion of a catalog-generation step in the inference
pipeline between the raw imaging and the final astrophysical analyses
is potentially lossy; the hard decisions of catalog making destroy
information.  Probability theory suggests that it may be less
lossy to pass forward not a catalog but a probabilisitic
description of all the catalogs that could be consistent with the
imaging---a posterior probability in (enormously large) space of possible catalogs.
Essentially, the creation of a catalog is an
attempt to answer the question,
``Given the image we have obtained, what objects are present in the field and
what are their properties?''.
This article represents an attempt at this ambitous goal in the specific
situation where the only objects in the field are stars or other point sources.

Beyond these philosophical concerns, there are practical issues:
Standard methods for constructing catalogs can have difficulty in some
challenging situations. For example, when multiple sources overlap partially
or completely, it can be difficult to determine how many sources are present,
and how much flux belongs to each source. In principle, the uncertainty about
the existence and properties of the stars can be significant and should be
propagated into any inferences about the stellar population. A Bayesian approach
that obtains the posterior distribution over catalog space (rather than a single
catalog estimate) has the potential to overcome these problems by deblending
objects when it is possible, and clearly indicating the uncertainty remaining
when it is not possible.

Sampling a posterior probability distribution for catalogs is a challenging
numerical task for a number of reasons. Firstly, the number $N$ of objects in the
image (and that should therefore be listed in the catalog) is itself unknown.
Secondly, if $N$ is large, then the parameter space
of positions and properties (flux, size, etc) of the objects is also large.
This can cause Markov Chain Monte Carlo (MCMC) algorithms difficulties -- they
may take a long time to converge to the target posterior distribution over
the space of catalogs. Thirdly, this problem is subject to the so-called
label-switching problem that is commonly encountered in mixture modeling
\citep[e.g.][]{label_switching}. Given any proposed catalog, another catalog that is
equally plausible is the catalog obtained by shuffling the entries of the first
catalog. This leads to a posterior distribution with $N!$ identical peaks in
parameter space. This can lead to difficulties with certain (generally very
effective) MCMC algorithms such as the affine-invariant stretch move
\citep{goodman, emcee}.

Bayesian object detection (as this problem is sometimes called) has been
implemented both inside and outside of astronomy
\citep[e.g.][]{object, 2003MNRAS.338..765H, 2011MNRAS.415.3462F}. However, the
\citet{2011MNRAS.415.3462F} approach makes the assumption of a
known number of objects $N$.
This assumption is required for the
{\tt MultiNest} sampler \citep{multinest} to be applicable.
Using the results from the known $N$
run, it is possible (under certain circumstances) to reconstruct what the
results would have been if an unknown-$N$ model had been used. However,
this will not work well in situations where there is significant confusion
(i.e. two or more sources overlap). What is really required is a variable
dimension model, where $N \in \{0, 1, 2, ... \}$ is an unknown quantity to be
inferred from the data \citep[e.g.][]{2003MNRAS.338..765H}.
The computational implementation of these models will require tools such as
reversible jump Markov Chain Monte Carlo \citep{rjmcmc}. Other statistical
methods have also been used to model crowded fields
\citep[e.g. maximum likelihood,][]{irwin}.
However, maximum likelihood is not completely appropriate in flexible models
because it may lead to overfitting. In this situation overfitting would result
in more stars being added to the model to explain small positive fluctuations
in the image which are actually due to noise. Various other techniques have
also been proposed in the literature \citep[e.g.][]{2002MNRAS.335...73M, 2007A&A...461..373M, 2009PASP..121.1429Z}.

@@

\section{Bayesian Inference}\label{sec:bayes}
To quantitatively model uncertainties and transform noise in observed data
into uncertainties in parameters of interest, Bayesian Inference is the appropriate
framework \citep{cox, jaynes, caticha, mackay}. Suppose there exist unknown parameters
(denoted
collectively by $\theta$) and we expect to obtain some data $x$. Our prior
state of knowledge about the parameters is modelled by a prior
probability distribution:
\begin{equation}
p(\theta).
\end{equation}
Note that this is a very concise notation \citep{hogg} and should be read
as ``the probability distribution for $\theta$''.
We also model how the parameters give rise to the data, via a generative model.
This is also known as a {\it sampling distribution}:
\begin{equation}
p(x|\theta).
\end{equation}
Despite the singular, the sampling distribution is actually a family of
probability distributions over the space of possible data sets, one probability
distribution for each possible value
of $\theta$. Note that the choice of
the sampling distribution is also an assumption about prior knowledge:
It models prior information about the fact that the data $x$ is connected to
the parameters $\theta$ in some way \citep{caticha}. Without this prior
knowledge, learning is impossible: there has to be some relationship between
the parameters and the data, otherwise it would be impossible to learn about
parameters by obtaining data.

When specific data $x^*$ are taken into account, our state of knowledge about $\theta$
gets updated from the prior distribution to the posterior distribution
via Bayes' rule:
\begin{eqnarray}
p(\theta|x=x^*) &\propto& p(\theta)p(x|\theta)|_{x=x^*} \\
&=& p(\theta)\mathcal{L}(\theta; x)
\end{eqnarray}
The term $p(x|\theta)|_{x=x^*} = \mathcal{L}(\theta; x)$
is the {\it likelihood function}, which is the
probability of obtaining the actual data set $x^*$ as a function of the
parameters. In the case that the sampling distribution is a probability
density function, the likelihood is the probability density function evaluated at the observed
data. This usually causes no problems, although one should be aware of the
Borel-Kolmogorov paradox \citep{jaynes}.
As suggested by the above notation, the likelihood function is obtained from the
sampling distribution with the actual data substituted in and is therefore
a function of the parameters only.

@@

For simplicity, we assume a uniform prior probability distribution for the
position of each star. This creates a strong preference for catalogs where
the stars are uniformly scattered across the image. Thus, this model is
appropriate for small patches of sky where the density of stars is approximately
uniform. In other scenarios, such as images of stellar clusters, it is possible
to parameterize the spatial distribution of the stars in a similar way to how
we have parameterized the luminosity function.

\subsection{The Sampling Distribution}
The sampling distribution is a probabilistic model for the process that
generates the data; it describes the probability distribution we would use
to predict the data if we happened to know the true catalog. In our case,
the data will be an $m \times n$ array of pixel intensities $I$:
\begin{eqnarray}
\{I_{ij}\}
\end{eqnarray}
where the central position of each pixel is:
\begin{eqnarray}
\{X_{ij}, Y_{ij}\}.
\end{eqnarray}
The image is assumed to be a noisy version of the true underlying intensity
field. Thus, we need
a prescription for simulating an image $\{I_{ij}\}$ from a catalog $\theta$:
\begin{eqnarray}
\theta = \left\{N, \beta, \left\{x_i, y_i\right\}_{i=1}^N,
\left\{f_i\right\}_{i=1}^N\right\}.
\end{eqnarray}
If we knew the true catalog, we could compute the
``mock image'' we would expect to see
in the absence of noise. This mock image (defined at every point on the sky)
is given by:
\begin{eqnarray}
\mathcal{M}(x, y) &=& \sum_{i=1}^N f_i \mathcal{P}(x - x_i, y - y_i)
\end{eqnarray}
where $\mathcal{P}$ is the pixel-convolved point spread function (PSF). The use
of a pixel-convolved PSF is computationally advantageous because the PSF need
only be evaluated at the center of each pixel, rather than integrated over each
pixel.

Throughout this paper we will
assume the pixel-convolved PSF is a weighted mixture of two concentric
circular
Gaussians (this model is also used by \citet{1987PASP...99..191S}) with widths
$s_1$ and $s_2$:
\begin{eqnarray}
\mathcal{P}(x, y) = \frac{w}{2\pi s_1^2}\exp
\left[
-\frac{1}{2s_1^2}\left(x^2 + y^2\right)
\right]
+ \frac{1-w}{2\pi s_2^2}\exp
\left[-\frac{1}{2s_2^2}\left(x^2 + y^2\right)
\right].
\end{eqnarray}
The pixellated observed image is assumed to be generated from the mock
image (evaluated at the center of each pixel) plus noise:
\begin{eqnarray}
I_{ij} &=& \mathcal{M}(X_{ij}, Y_{ij}) + \epsilon_{ij}
\end{eqnarray}
where the errors $\{\epsilon_{ij}\}$ are independent and normally distributed.
The variance of each pixel is determined by the brightness of the sky and the
brightness of the mock image in that pixel:
\begin{eqnarray}
\epsilon_{ij} \sim \mathcal{N}(0, \sigma_0^2 + \eta \mathcal{M}(X_{ij}, Y_{ij})).
\end{eqnarray}
where $\sigma_0$ is a constant noise level and $\eta$ is a coefficient allowing
the noise to be higher in brighter regions of the image. This dependence of the
noise variances on the model intensity arises as a result of the Poissonian
nature of photon counts, but allows for the fact that a ``sky'' background may
have already been subtracted from the image.
This parameterisation has been used by \citet{2011MNRAS.412.2521B} and is an
alternative to the common practice of producing a ``variance map'' from the
image data that is then assumed to be known.

\subsection{The Prior Distribution}
The prior distribution for the number of stars $N$ is assigned to be uniform
between 0 and some maximum number $N_{\rm max}$. The extent of the image is
assumed to be from $x=x_{\rm min}$ to $x=x_{\rm max}=x_{\rm min} + x_{\rm range}$ and from
$y=y_{\rm min}$ to $y=y_{\rm max}=y_{\rm min} + y_{\rm range}$ in arbitrary units, and the
positions of the stars are assigned independent uniform priors:
\begin{eqnarray}
x_i &\sim& \textnormal{Uniform}(x_{\rm min} - 0.1x_{\rm range}, x_{\rm max} + 0.1x_{\rm range}) \\
y_i &\sim& \textnormal{Uniform}(y_{\rm min} - 0.1y_{\rm range}, y_{\rm max} + 0.1y_{\rm range})
\end{eqnarray}
The stars are allowed to be slightly outside of the observed image because the
PSF can scatter light from these stars into the image.

For the purposes of this paper, we model the luminosity function as a
luminosity function is a broken power-law with four
free parameters:
\begin{eqnarray}
\beta = \{h_1, h_2, \alpha_1, \alpha_2\}.
\end{eqnarray}
where $h_1$ is a lower flux limit, $h_2$ is a break-point, $\alpha_1$ is
the slope of the distribution between $h_1$ and $h_2$, and $\alpha_2$ is the
slope of the distribution above $h_2$. For mathematical details on the broken power-law
model, see Appendix~\ref{power_law}. While the broken power-law is likely to be
unrealistic in many cases, it is a reasonably flexible distribution and this is
sufficient for demonstrating the properties of out method.

The prior distribution on $h_1$, $h_2$,
$\alpha_1$, and $\alpha_2$ is assigned to be:
\begin{eqnarray}
\ln h_1 &\sim& \textnormal{Uniform}(\ln(10^{-3}), \ln(10^3)) \\
\ln h_2 &\sim& \textnormal{Uniform}(\ln(h_1), \ln(h_1) + 2.3) \\
\alpha_1 &\sim& \textnormal{Uniform}(1, 5) \\
\alpha_2 &\sim& \textnormal{Uniform}(1, 5).
\end{eqnarray}
These priors express vague prior knowledge about $\alpha_1$ and $\alpha_2$ in
addition to vague prior knowledge about $h_1$ and $h_2$ apart from the fact
that the flux units are not extreme and that $h_2$ should be no more than an
order of magnitude greater than $h_1$.

@@

The priors for the PSF parameters and the noise parameters were assigned to be:
\begin{eqnarray}
\ln s_1 &\sim& \textnormal{Uniform}(\ln(0.3L), \ln(30L)) \\
\ln s_2 &\sim& \textnormal{Uniform}(\ln(s_1), \ln(s_1) + 2.3) \\
w &\sim& \textnormal{Uniform}(0, 1)\\
\ln \sigma_0 &\sim& \textnormal{Uniform}(\ln(10^{-3}), \ln(10^{3})) \\
\ln \eta &\sim& \textnormal{Uniform}(\ln(10^{-3}), \ln(10^{3}))
\end{eqnarray}
where $L = (x_{\rm max} - x_{\rm min})/n$ is the width of a pixel. These priors
describe vague prior knowledge about the overall scale of the PSF except that the
wider component is less than 10 times as wide as the narrow component, as well
as the knowledge that the noise variance is not extreme relative to the fluxes
of the stars.

@@

\section{Simulated Data}\label{sec:simulated_data}
In order to test our approach, we applied the method to two illustrative
simulated images generated from
the above model (Figure~\ref{fig:simulated_data}). The purpose of this
experiment was to test the computational
feasibility of the model, as well as to compare the inferences from the model
with those from more standard techniques.

The true parameter values for the two simulated data sets are listed in
Table~\ref{tab:truth}. The broken power-law parameter values were chosen so that
roughly half of the stars' fluxes were below and above the break-point
respectively. Figure~\ref{fig:powerlaw} in Appendix~\ref{power_law} also shows the true flux distribution
used for the simulated images. Each of the images
is $100 \times 100$ pixels in extent and covers a range from $-1$ to $1$ in
arbitrary units for both the $x$ and $y$ axes. The first image contains
100 stars (including stars just outside of the image; there are 63 stars whose central
positions lie within the image) and the second
image contains $\sim$1000 stars (699 of which are positioned within the image).

\begin{table}[ht!]\footnotesize
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Parameter & Value (Test Case 1) & Value (Test Case 2)\\
\hline
$N$ & 100 & 1000\\
$h_1$ & 0.3 & 0.3\\
$h_2$ & 0.6 & 0.6\\
$\alpha_1$ & 1.1 & 1.1\\
$\alpha_2$ & 2 & 2\\
\hline
$\sigma_0$ & 10 & 10\\
$\eta$ & 10 & 10 \\
$s_1$ & 0.02 & 0.02\\
$s_2$ & 0.1 & 0.1\\
$w$ & 0.5 & 0.5\\
\hline
\end{tabular}
\end{center}
\caption{True parameter values used to generate the simulated data. The only
difference between the two images is that Test Case 2 contains more stars than
Test Case 1.
\label{tab:truth}}
\end{table}

@@

\subsection{Test Case 1}
Test Case 1 was run with the DNS algorithm and usable results were obtained
within about an hour on a modern desktop PC. The inferences on the parameters
$N$, $h_1$, $h_2$, $\alpha_1$, and $\alpha_2$ are shown in
Figure~\ref{fig:results1}. The number of stars is correctly inferred, and the
posterior distributions for the other parameters comfortably contain the true
input values. Note that the uncertainty in $h_2$, $\alpha_1$, and $\alpha_2$ is
quite large. This is because the broken power-law model
(Figure~\ref{fig:powerlaw}) does not change drastically in shape as the
parameters are varied. Therefore, a large number of stars would be required to
tightly constrain these parameters.

@@

\subsection{Test Case 2}
Test Case 2 is more challenging than Test Case 1 because the image contains
more stars. This increases the size of the computational task in a number of
ways: firstly, there will be more unknown parameters to infer, so any MCMC
algorithm will require more iterations in order to converge to the posterior
distribution. Secondly, the time taken
to compute the predicted image from a proposed catalog (in order to evaluate the
likelihood) is longer because of the larger number of stars. Using DNS, some
samples from the posterior distribution can be obtained in about a day on a
modern PC.

@@

\begin{table}[ht!]\footnotesize
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Parameter & Proposal & Notes\\
\hline
$N$ & $N \to N + \delta_N$ & Generate $\delta_N$ new stars from
$p(x, y, f|\beta)$.\\
$N$ & $N \to N - \delta_N$ & Remove $\delta_N$ stars, chosen at random\\
$\beta$ & $\beta \to \beta + \delta_\beta$ & Transform
stars' fluxes correspondingly\\
$\beta$ & $\beta \to \beta + \delta_\beta$ & Fix stars' fluxes, put extra
term in acceptance probability \\
$(x_i,y_i)$ & $(x_i,y_i) \to (x_i,y_i)+(\delta_x, \delta_y)$ & Can move $>1$ star
in a single step \\
$f$ & $f \to f + \delta_f$ & Can move $>1$ stars' fluxes in a single step\\
\hline
\end{tabular}
\end{center}
\caption{All $\delta$ parameters are drawn from multi-scale distibutions such
that the largest steps are of order the prior width, and the smallest steps
are of order $10^{-6}$ times the prior width.\label{tab:proposals}}
\end{table}
